---
title: "Check the DEGs of targeted genes by GSFA vs other methods"
author: "Kaixuan Luo"
date: "2022-09-01"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
editor_options:
  chunk_output_type: console
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(results = "hold", fig.align = "center", message=FALSE, warning=FALSE)
```

## Load necessary packages and data

```{r}
library(Seurat)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)
library(WebGestaltR)
library(GSFA)
library(DT)
source("code/plotting_functions.R")
```

Set directories
```{r dir}
res_dir <- "/project2/xinhe/kevinluo/GSFA/DEGs_by_expression_level"
dir.create(res_dir, recursive = TRUE, showWarnings = FALSE)
```

# LUHMES data

Load GSFA result
```{r}
data_folder <- "/project2/xinhe/yifan/Factor_analysis/LUHMES/"
fit <- readRDS(paste0(data_folder,
                      "gsfa_output_detect_01/use_negctrl/All.gibbs_obj_k20.svd_negctrl.seed_14314.light.rds"))
gibbs_PM <- fit$posterior_means
lfsr_mat <- fit$lfsr[, -ncol(fit$lfsr)]
total_effect <- fit$total_effect[, -ncol(fit$total_effect)]
KO_names <- colnames(lfsr_mat)
guides <- KO_names[KO_names!="Nontargeting"]
```

Load the mapping from gene name to ENSEMBL ID for all 6k genes used in GSFA
```{r}
feature.names <- data.frame(fread(paste0(data_folder, "GSE142078_raw/GSM4219576_Run2_genes.tsv.gz"),
                                  header = FALSE), stringsAsFactors = FALSE)
genes_df <- feature.names[match(rownames(lfsr_mat), feature.names$V1), ]
names(genes_df) <- c("ID", "Name")
```


Check the gene loadings to see if the targeted genes in any of the factors P(F) > 0.95?
```{r}
F_pm <- gibbs_PM$F_pm
guide_genes_df <- feature.names[match(guides, feature.names$V2), ]
names(guide_genes_df) <- c("ID", "Name")

F_pm_guides <- F_pm[match(guide_genes_df$ID, rownames(F_pm)), ]
rownames(F_pm_guides) <- guides

F_pm_guides_highpip <- ifelse(F_pm_guides > 0.95, 1, 0)
F_pm_guides_highpip[is.na(F_pm_guides_highpip)] <- 0
gsfa_num_highpip_factors <- rowSums(F_pm_guides_highpip)
```


## DEGs for targeted genes

GSFA result
```{r gsfa_LUHMES}
effect_mat <- total_effect
rownames(effect_mat) <- rownames(lfsr_mat)
colnames(effect_mat) <- colnames(lfsr_mat)

gsfa_deg_guides.df <- data.frame()
for(guide in guides){
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% rownames(lfsr_mat)){
    gsfa_deg_guides.df <- rbind(gsfa_deg_guides.df, 
                 data.frame(guide = guide, lfsr = lfsr_mat[guide_geneID, guide], effect = effect_mat[guide_geneID, guide]))
  }else{
    gsfa_deg_guides.df <- rbind(gsfa_deg_guides.df, 
                          data.frame(guide = guide, lfsr = NA, effect = NA))
  }
}

gsfa_deg_guides.df$sig_DE <- ifelse(gsfa_deg_guides.df$lfsr < 0.05, "DE", "")
```


MAST
```{r mast_LUHMES}
mast_deg_guides.df <- data.frame()
for (guide in guides){
  fname <- paste0(data_folder, "processed_data/MAST/dev_top6k_negctrl/gRNA_", 
                  guide, ".dev_res_top6k.vs_negctrl.rds")
  mast_df <- readRDS(fname)
  mast_df$geneID <- rownames(mast_df)
  mast_df <- mast_df %>% dplyr::rename(FDR = fdr, PValue = pval)
  
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% rownames(mast_df)){
    mast_deg_guides.df <- rbind(mast_deg_guides.df, 
                 data.frame(guide = guide, fdr = mast_df[guide_geneID, "FDR"], logFC = mast_df[guide_geneID, "logFC"]))
  }else{
    mast_deg_guides.df <- rbind(mast_deg_guides.df, 
                          data.frame(guide = guide, fdr = NA, logFC = NA))
  }
}

mast_deg_guides.df$fdr <- round(mast_deg_guides.df$fdr, 2)
mast_deg_guides.df$logFC <- round(mast_deg_guides.df$logFC, 2)
mast_deg_guides.df$sig_DE <- ifelse(mast_deg_guides.df$fdr < 0.05, "DE", "")
```

DESeq2
```{r deseq_LUHMES}
deseq_deg_guides.df <- data.frame()

for (guide in guides){
  fname <- paste0(data_folder, "processed_data/DESeq2/dev_top6k_negctrl/gRNA_", 
                  guide, ".dev_res_top6k.vs_negctrl.rds")
  res <- readRDS(fname)
  res <- as.data.frame(res@listData, row.names = res@rownames)
  res$geneID <- rownames(res)
  deseq_df <- res %>% dplyr::rename(FDR = padj, PValue = pvalue)
  
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% rownames(deseq_df)){
    deseq_deg_guides.df <- rbind(deseq_deg_guides.df, 
                 data.frame(guide = guide, fdr = deseq_df[guide_geneID, "FDR"], logFC = deseq_df[guide_geneID, "log2FoldChange"]))
  }else{
    deseq_deg_guides.df <- rbind(deseq_deg_guides.df, 
                          data.frame(guide = guide, fdr = NA, logFC = NA))
  }
}


deseq_deg_guides.df$fdr <- round(deseq_deg_guides.df$fdr, 2)
deseq_deg_guides.df$logFC <- round(deseq_deg_guides.df$logFC, 2)
deseq_deg_guides.df$sig_DE <- ifelse(deseq_deg_guides.df$fdr < 0.05, "DE", "")
```

scMAGeCK
```{r scmageck_LUHMES}
scmageck_res <- readRDS(paste0(data_folder, "scmageck/scmageck_lr.LUHMES.dev_res_top_6k.rds"))
colnames(scmageck_res$fdr)[colnames(scmageck_res$fdr) == "NegCtrl"] <- "Nontargeting"
colnames(scmageck_res$score)[colnames(scmageck_res$score) == "NegCtrl"] <- "Nontargeting"

scmageck_deg_guides.df <- data.frame()
for (guide in guides){
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% rownames(scmageck_res$fdr)){
    scmageck_deg_guides.df <- rbind(scmageck_deg_guides.df, 
                 data.frame(guide = guide, fdr = scmageck_res$fdr[guide_geneID, guide], score = scmageck_res$score[guide_geneID, guide]))
  }else{
    scmageck_deg_guides.df <- rbind(scmageck_deg_guides.df, 
                          data.frame(guide = guide, fdr = NA, score = NA))
  }
}

scmageck_deg_guides.df$fdr <- round(scmageck_deg_guides.df$fdr, 2)
scmageck_deg_guides.df$score <- round(scmageck_deg_guides.df$score, 2)
scmageck_deg_guides.df$sig_DE <- ifelse(scmageck_deg_guides.df$fdr < 0.05, "DE", "")
```

SCEPTRE
```{r sceptre_LUHMES}
sceptre_res <- readRDS("/project2/xinhe/kevinluo/GSFA/sceptre_analysis/LUHMES_cropseq_data/sceptre_output/sceptre.result.rds")
sceptre_count_df <- data.frame(matrix(nrow = length(guides), ncol = 2))
colnames(sceptre_count_df) <- c("target", "num_DEG")

sceptre_deg_guides.df <- data.frame()
for (guide in guides){
  curr_sceptre_res <- sceptre_res %>% filter(gRNA_id == guide)
  curr_sceptre_res$fdr <- p.adjust(curr_sceptre_res$p_value, method = "fdr")
  
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% curr_sceptre_res$gene_id){
    sceptre_deg_guides.df <- rbind(sceptre_deg_guides.df, 
                 data.frame(guide = guide, fdr = curr_sceptre_res$fdr[curr_sceptre_res$gene_id == guide_geneID], 
                            z_value = curr_sceptre_res$z_value[curr_sceptre_res$gene_id == guide_geneID]))
  }else{
    sceptre_deg_guides.df <- rbind(sceptre_deg_guides.df, 
                          data.frame(guide = guide, fdr = NA, z_value = NA))
  }
}

sceptre_deg_guides.df$fdr <- round(sceptre_deg_guides.df$fdr, 2)
sceptre_deg_guides.df$z_value <- round(sceptre_deg_guides.df$z_value, 2)
sceptre_deg_guides.df$sig_DE <- ifelse(sceptre_deg_guides.df$fdr < 0.05, "DE", "")

```

Compare across methods
```{r deg_comparison_LUHMES}
dge_guides_comparison_df <- data.frame(Perturbation = guides,
                                       num_GSFA_factors = gsfa_num_highpip_factors, 
                                       GSFA = gsfa_deg_guides.df$sig_DE,
                                       scMAGeCK = scmageck_deg_guides.df$sig_DE,
                                       DESeq2 = deseq_deg_guides.df$sig_DE,
                                       MAST = mast_deg_guides.df$sig_DE,
                                       SCEPTRE = sceptre_deg_guides.df$sig_DE)

DT::datatable(dge_guides_comparison_df, 
              rownames = FALSE, 
              options = list(pageLength = length(guides), 
                             columnDefs = list(list(className = 'dt-center', targets = 0:4))))
```

# T cells data

Load GSFA result
```{r}
data_folder <- "/project2/xinhe/yifan/Factor_analysis/Stimulated_T_Cells/"
fit <- readRDS(paste0(data_folder,
                      "gsfa_output_detect_01/all_uncorrected_by_group.use_negctrl/All.gibbs_obj_k20.svd_negctrl.restart.light.rds"))
gibbs_PM <- fit$posterior_means
lfsr_mat1 <- fit$lfsr1[, -ncol(fit$lfsr1)]
lfsr_mat0 <- fit$lfsr0[, -ncol(fit$lfsr0)]
total_effect1 <- fit$total_effect1[, -ncol(fit$total_effect1)]
total_effect0 <- fit$total_effect0[, -ncol(fit$total_effect0)]
KO_names <- colnames(lfsr_mat1)
guides <- KO_names[KO_names!="NonTarget"]
```

Load the mapping from gene name to ENSEMBL ID for all 6k genes used in GSFA.

```{r}
feature.names <- data.frame(fread(paste0(data_folder, "GSE119450_RAW/D1N/genes.tsv"),
                                  header = FALSE), stringsAsFactors = FALSE)
genes_df <- feature.names[match(rownames(lfsr_mat1), feature.names$V1), ]
names(genes_df) <- c("ID", "Name")
```

Check the gene loadings to see if the targeted genes in any of the factors P(F) > 0.95?
```{r}
F_pm <- gibbs_PM$F_pm
guide_genes_df <- feature.names[match(guides, feature.names$V2), ]
names(guide_genes_df) <- c("ID", "Name")

F_pm_guides <- F_pm[match(guide_genes_df$ID, rownames(F_pm)), ]
rownames(F_pm_guides) <- guides

F_pm_guides_highpip <- ifelse(F_pm_guides > 0.95, 1, 0)
F_pm_guides_highpip[is.na(F_pm_guides_highpip)] <- 0
gsfa_num_highpip_factors <- rowSums(F_pm_guides_highpip)
```


## DEGs for targeted genes

GSFA result
```{r gsfa_t_cells}
lfsr_mat <- lfsr_mat1
effect_mat <- total_effect1
rownames(effect_mat) <- rownames(lfsr_mat)
colnames(effect_mat) <- colnames(lfsr_mat)

gsfa_deg_guides.df <- data.frame()
for(guide in guides){
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% rownames(lfsr_mat)){
    gsfa_deg_guides.df <- rbind(gsfa_deg_guides.df, 
                 data.frame(guide = guide, lfsr = lfsr_mat[guide_geneID, guide], effect = effect_mat[guide_geneID, guide]))
  }else{
    gsfa_deg_guides.df <- rbind(gsfa_deg_guides.df, 
                          data.frame(guide = guide, lfsr = NA, effect = NA))
  }
}

gsfa_deg_guides.df$sig_DE <- ifelse(gsfa_deg_guides.df$lfsr < 0.05, "DE", "")
```

MAST
```{r mast_t_cells}
mast_deg_guides.df <- data.frame()
for (guide in guides){
  fname <- paste0(data_folder, "processed_data/MAST/all_by_stim_negctrl/gRNA_", 
                  guide, ".dev_res_top6k.vs_negctrl.rds")
  mast_df <- readRDS(fname)
  mast_df$geneID <- rownames(mast_df)
  mast_df <- mast_df %>% dplyr::rename(FDR = fdr, PValue = pval)
  
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% rownames(mast_df)){
    mast_deg_guides.df <- rbind(mast_deg_guides.df, 
                 data.frame(guide = guide, fdr = mast_df[guide_geneID, "FDR"], logFC = mast_df[guide_geneID, "logFC"]))
  }else{
    mast_deg_guides.df <- rbind(mast_deg_guides.df, 
                          data.frame(guide = guide, fdr = NA, logFC = NA))
  }
}

mast_deg_guides.df$fdr <- round(mast_deg_guides.df$fdr, 2)
mast_deg_guides.df$logFC <- round(mast_deg_guides.df$logFC, 2)
mast_deg_guides.df$sig_DE <- ifelse(mast_deg_guides.df$fdr < 0.05, "DE", "")
```

DESeq2
```{r deseq_t_cells}
deseq_deg_guides.df <- data.frame()

for (guide in guides){
  fname <- paste0(data_folder, "processed_data/DESeq2/all_by_stim_negctrl/gRNA_", 
                  guide, ".dev_res_top6k.vs_negctrl.rds")
  res <- readRDS(fname)
  res <- as.data.frame(res@listData, row.names = res@rownames)
  res$geneID <- rownames(res)
  deseq_df <- res %>% dplyr::rename(FDR = padj, PValue = pvalue)
  
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% rownames(deseq_df)){
    deseq_deg_guides.df <- rbind(deseq_deg_guides.df, 
                 data.frame(guide = guide, fdr = deseq_df[guide_geneID, "FDR"], logFC = deseq_df[guide_geneID, "log2FoldChange"]))
  }else{
    deseq_deg_guides.df <- rbind(deseq_deg_guides.df, 
                          data.frame(guide = guide, fdr = NA, logFC = NA))
  }
}


deseq_deg_guides.df$fdr <- round(deseq_deg_guides.df$fdr, 2)
deseq_deg_guides.df$logFC <- round(deseq_deg_guides.df$logFC, 2)
deseq_deg_guides.df$sig_DE <- ifelse(deseq_deg_guides.df$fdr < 0.05, "DE", "")

```

scMAGeCK
```{r scmageck_t_cells}
scmageck_res <- readRDS(paste0(data_folder, "scmageck/scmageck_lr.TCells_stim.dev_res_top_6k.rds"))
colnames(scmageck_res$fdr)[colnames(scmageck_res$fdr) == "NegCtrl"] <- "NonTarget"
colnames(scmageck_res$score)[colnames(scmageck_res$score) == "NegCtrl"] <- "NonTarget"

scmageck_deg_guides.df <- data.frame()
for (guide in guides){
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% rownames(scmageck_res$fdr)){
    scmageck_deg_guides.df <- rbind(scmageck_deg_guides.df, 
                 data.frame(guide = guide, fdr = scmageck_res$fdr[guide_geneID, guide], score = scmageck_res$score[guide_geneID, guide]))
  }else{
    scmageck_deg_guides.df <- rbind(scmageck_deg_guides.df, 
                          data.frame(guide = guide, fdr = NA, score = NA))
  }
}

scmageck_deg_guides.df$fdr <- round(scmageck_deg_guides.df$fdr, 2)
scmageck_deg_guides.df$score <- round(scmageck_deg_guides.df$score, 2)
scmageck_deg_guides.df$sig_DE <- ifelse(scmageck_deg_guides.df$fdr < 0.05, "DE", "")
```

SCEPTRE
```{r sceptre_t_cells}
sceptre_res <- readRDS("/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data/sceptre_output/sceptre.result.rds")
sceptre_count_df <- data.frame(matrix(nrow = length(guides), ncol = 2))
colnames(sceptre_count_df) <- c("target", "num_DEG")

sceptre_deg_guides.df <- data.frame()
for (guide in guides){
  curr_sceptre_res <- sceptre_res %>% filter(gRNA_id == guide)
  curr_sceptre_res$fdr <- p.adjust(curr_sceptre_res$p_value, method = "fdr")
  
  guide_geneID <- feature.names[which(feature.names$V2 == guide),]$V1
  if(guide_geneID %in% curr_sceptre_res$gene_id){
    sceptre_deg_guides.df <- rbind(sceptre_deg_guides.df, 
                 data.frame(guide = guide, fdr = curr_sceptre_res$fdr[curr_sceptre_res$gene_id == guide_geneID], 
                            z_value = curr_sceptre_res$z_value[curr_sceptre_res$gene_id == guide_geneID]))
  }else{
    sceptre_deg_guides.df <- rbind(sceptre_deg_guides.df, 
                          data.frame(guide = guide, fdr = NA, z_value = NA))
  }
}

sceptre_deg_guides.df$fdr <- round(sceptre_deg_guides.df$fdr, 2)
sceptre_deg_guides.df$z_value <- round(sceptre_deg_guides.df$z_value, 2)
sceptre_deg_guides.df$sig_DE <- ifelse(sceptre_deg_guides.df$fdr < 0.05, "DE", "")

```

Compare across methods
```{r deg_comparison_t_cells}
dge_guides_comparison_df <- data.frame(Perturbation = guides,
                                       num_GSFA_factors = gsfa_num_highpip_factors, 
                                       GSFA = gsfa_deg_guides.df$sig_DE,
                                       scMAGeCK = scmageck_deg_guides.df$sig_DE,
                                       DESeq2 = deseq_deg_guides.df$sig_DE,
                                       MAST = mast_deg_guides.df$sig_DE,
                                       SCEPTRE = sceptre_deg_guides.df$sig_DE)


DT::datatable(dge_guides_comparison_df, 
              rownames = FALSE, 
              options = list(pageLength = length(guides), 
                             columnDefs = list(list(className = 'dt-center', targets = 0:4))))
```
