---
title: "SCEPTRE analysis on LUHMES CROP-seq data"
author: "Kaixuan Luo"
date: "2022-05-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

## About the data sets

CROP-seq datasets: `/project2/xinhe/yifan/Factor_analysis/shared_data/`
The data are Seurat objects, with raw gene counts stored in obj@assays$RNA@counts, and cell meta data stored in obj@meta.data.
Normalized and scaled data used for GSFA are stored in obj@assays$RNA@scale.data , the rownames of which are the 6k genes used for GSFA.

## Analysis

Scripts for running the analysis:

* [R script](https://github.com/kevinlkx/GSFA_analysis/tree/main/code/sceptre_analysis.R).
* [sbatch script](https://github.com/kevinlkx/GSFA_analysis/tree/main/code/run_sceptre_cropseq_data.sbatch).

```{bash, eval=FALSE}
cd /project2/xinhe/kevinluo/GSFA/log

sbatch --mem=50G --cpus-per-task=10 ~/projects/GSFA_analysis/code/run_sceptre_cropseq_data.sbatch /project2/xinhe/kevinluo/GSFA/sceptre_analysis/LUHMES_cropseq_data
```

Load packages
```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(tidyverse))
library(cowplot)
library(Matrix)
library(sceptre)
library(Seurat)
```

### Prepare input data for SCEPTRE

Load the LUHMES data
```{r load-data}
LUHMES_data <- readRDS('/project2/xinhe/yifan/Factor_analysis/shared_data/LUHMES_cropseq_data_seurat.rds')
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/LUHMES_cropseq_data'
```

We first prepare three separate data objects required to run SCEPTRE: 
the gene expression matrix, the perturbation matrix, and the covariate matrix.

1. Gene expression matrices
Gene expression (gene-by-cell) raw count matrix
```{r exp-matrix}
gene_matrix <- LUHMES_data@assays$RNA@counts
# gene-by-cell expression matrix
gene_matrix[1:10, 1:3]
dim(gene_matrix)
```

2. Cell meta data and covariate matrix
```{r cell-covariate-matrix}
metadata <- LUHMES_data@meta.data
metadata[1:5, ]
covariate_matrix <- metadata[,c('orig.ident', 'nCount_RNA', 'nFeature_RNA', 'percent_mt')]
covariate_matrix[1:5,]
dim(covariate_matrix)
```

3. Perturbation matrix (a binary matrix of perturbations,  rows are gRNA groups and columns are cell barcodes)
```{r perturbation-matrix}
combined_perturbation_matrix <- t(metadata[,4:18])

dim(combined_perturbation_matrix)

combined_perturbation_matrix[1:10,1:3]

range(combined_perturbation_matrix)
```

Specify the gene-gRNA group pairs to test for association

We include the 6k genes used for GSFA in this analysis
```{r select-genes}
# Normalized and scaled data used for GSFA, the rownames of which are the 6k genes used for GSFA
scaled_gene_matrix <- LUHMES_data@assays$RNA@scale.data
dim(scaled_gene_matrix)
selected_gene_id <- rownames(scaled_gene_matrix)
all(selected_gene_id %in% rownames(gene_matrix))
```

```{r select-pairs, eval=FALSE}
gRNA_group <- rownames(combined_perturbation_matrix)
pairs <- expand.grid(selected_gene_id, gRNA_group)
gene_gRNA_group_pairs <- data.frame(gene_id = pairs$Var1, gRNA_group = pairs$Var2, pair_type = "candidate")
gene_gRNA_group_pairs[gene_gRNA_group_pairs$gRNA_group == "Nontargeting", "pair_type"] <- "negative_control"
table(gene_gRNA_group_pairs$pair_type)
table(gene_gRNA_group_pairs$gRNA_group)
dim(gene_gRNA_group_pairs)
```

```{r, eval=FALSE}
save(list = c("gene_matrix", "combined_perturbation_matrix", "covariate_matrix"),
     file = file.path(datadir, 'data.matrices.RData'))
saveRDS(gene_gRNA_group_pairs, file.path(datadir, "gene.gRNA.group.pairs.rds"))
```

### Run SCEPTRE
```{r load-input-data}
load(file.path(datadir, "data.matrices.RData"))
gene_gRNA_group_pairs <- readRDS(file.path(datadir, "gene.gRNA.group.pairs.rds"))

cat(sprintf('Dimenstion of gene expression matrix: %d rows %d columns.\n', nrow(gene_matrix), ncol(gene_matrix)))
cat(sprintf('Dimenstion of combined perturbation matrix: %d rows %d columns.\n', nrow(combined_perturbation_matrix), ncol(combined_perturbation_matrix)))
cat(sprintf('Dimenstion of covariate matrix: %d rows %d columns.\n', nrow(covariate_matrix), ncol(covariate_matrix)))
cat(sprintf('Dimenstion of gene gRNA-group pairs: %d rows %d columns.\n', nrow(gene_gRNA_group_pairs), ncol(gene_gRNA_group_pairs)))

table(gene_gRNA_group_pairs$pair_type)
```

```{r run-sceptre, eval=FALSE}
result <- run_sceptre_high_moi(gene_matrix = gene_matrix,
                               combined_perturbation_matrix = combined_perturbation_matrix,
                               covariate_matrix = covariate_matrix,
                               gene_gRNA_group_pairs = gene_gRNA_group_pairs,
                               side = "both",
                               storage_dir = outdir, 
                               full_output = TRUE)

outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/LUHMES_cropseq_data/sceptre_output'
saveRDS(result, file.path(outdir, 'sceptre.result.rds'))
```

### Analyze the results
```{r load-result}
outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/LUHMES_cropseq_data/sceptre_output'
result <- readRDS(file.path(outdir, 'sceptre.result.rds'))
head(result, 10)
```

Negative control pairs
```{r neg-control}
neg_control_p_vals <- result %>% filter(pair_type == "negative_control") %>% pull(p_value)
qq_plot <- make_qq_plot(neg_control_p_vals)
plot(qq_plot)
```

Candidate pairs
We extract the p-values corresponding to the candidate pairs and apply a Benjamini-Hochberg (BH) correction to adjust for multiple testing.
```{r fdr}
candidate_pair_results <- result %>% filter(pair_type == "candidate")
candidate_pair_results_p_adj <- candidate_pair_results %>%
  mutate(p_val_adj = p.adjust(p_value, method = "BH"))
head(candidate_pair_results_p_adj)
```

We call pairs with an adjusted p-value of less or equal than 0.1 significant; the discovery set (i.e., the set of significant pairs) has a false discovery rate (FDR) of 10%.
```{r select-sig-pairs}
discovery_set <- candidate_pair_results_p_adj %>% filter(p_val_adj <= 0.1)
head(discovery_set)
```

```{r}
saveRDS(candidate_pair_results_p_adj, file.path(outdir, 'sceptre.candidate.pair.results.rds'))
saveRDS(discovery_set, file.path(outdir, 'sceptre.discovery.set.results.rds'))
```


