---
title: "Run SCEPTRE on CD8+ T Cell (stimulated and unstimulated) CROP-seq data"
author: "Kaixuan Luo"
date: "2022-05-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

## About the data sets

CROP-seq datasets: `/project2/xinhe/yifan/Factor_analysis/shared_data/`
The data are Seurat objects, with raw gene counts stored in obj@assays$RNA@counts, and cell meta data stored in obj@meta.data.
Normalized and scaled data used for GSFA are stored in obj@assays$RNA@scale.data , the rownames of which are the 6k genes used for GSFA.

## Analysis

Scripts for running the analysis:

* [R script](../code/sceptre_analysis.R).
* [sbatch script](../code/run_sceptre_cropseq_data.sbatch).

```{bash, eval=FALSE}
cd /project2/xinhe/kevinluo/GSFA/log

sbatch --mem=50G --cpus-per-task=10 ~/projects/GSFA_analysis/code/run_sceptre_cropseq_data.sbatch /project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data

sbatch --mem=50G --cpus-per-task=10 ~/projects/GSFA_analysis/code/run_sceptre_cropseq_data.sbatch /project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/unsimulated_data
```

Load packages
```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(tidyverse))
library(cowplot)
library(Matrix)
library(sceptre)
library(Seurat)
```

### Prepare input data for SCEPTRE

Load the T Cells CROP-seq data
```{r load-data}
TCells_data <- readRDS('/project2/xinhe/yifan/Factor_analysis/shared_data/TCells_cropseq_data_seurat.rds')
```

We first prepare three separate data objects required to run SCEPTRE: 
the gene expression matrix, the perturbation matrix, and the covariate matrix.

Separate stimulated and unstimulated cells into two data sets, and run those separately.
```{r cell-covariate-matrix}
metadata <- TCells_data@meta.data
metadata[1:5, ]

table(metadata$orig.ident)

stimulated_cells <- rownames(metadata)[which(endsWith(metadata$orig.ident, "S"))]
cat(length(stimulated_cells), "stimulated cells. \n")
unstimulated_cells <- rownames(metadata)[which(endsWith(metadata$orig.ident, "N"))]
cat(length(unstimulated_cells), "unstimulated cells. \n")
```

#### Stimulated T cells

We first prepare three separate data objects required to run SCEPTRE: 
the gene expression matrix, the perturbation matrix, and the covariate matrix.

1. Gene expression matrices
Gene expression (gene-by-cell) raw count matrix
```{r stimulated-exp-matrix}
gene_matrix <- TCells_data@assays$RNA@counts[, stimulated_cells]
# gene-by-cell expression matrix
gene_matrix[1:5, 1:3]
dim(gene_matrix)
```

2. Covariate matrix
```{r stimulated-covariate-matrix}
covariate_matrix <- metadata[stimulated_cells, c('nCount_RNA', 'nFeature_RNA', 'percent_mt', 'gRNA_umi_count')]
covariate_matrix[1:5,]
dim(covariate_matrix)
```

3. Perturbation matrix (a binary matrix of perturbations,  rows are gRNA groups and columns are cell barcodes)
```{r stimulated-perturbation-matrix}
combined_perturbation_matrix <- t(metadata[stimulated_cells, 4:24])
combined_perturbation_matrix[1:5,1:3]
dim(combined_perturbation_matrix)
range(combined_perturbation_matrix)
```

```{r, eval=FALSE}
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data'
save(list = c("gene_matrix", "combined_perturbation_matrix", "covariate_matrix"),
     file = file.path(datadir, 'data.matrices.RData'))
```


#### Unstimulated T cells

1. Gene expression matrices
Gene expression (gene-by-cell) raw count matrix
```{r unstimulated-exp-matrix}
gene_matrix <- TCells_data@assays$RNA@counts[, unstimulated_cells]
# gene-by-cell expression matrix
gene_matrix[1:5, 1:3]
dim(gene_matrix)
```

2. Covariate matrix
```{r unstimulated-covariate-matrix}
covariate_matrix <- metadata[unstimulated_cells, c('nCount_RNA', 'nFeature_RNA', 'percent_mt', 'gRNA_umi_count')]
covariate_matrix[1:5,]
dim(covariate_matrix)
```

3. Perturbation matrix (a binary matrix of perturbations,  rows are gRNA groups and columns are cell barcodes)
```{r unstimulated-perturbation-matrix}
combined_perturbation_matrix <- t(metadata[unstimulated_cells, 4:24])
combined_perturbation_matrix[1:5,1:3]
dim(combined_perturbation_matrix)
range(combined_perturbation_matrix)
```


```{r, eval=FALSE}
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/unsimulated_data'
save(list = c("gene_matrix", "combined_perturbation_matrix", "covariate_matrix"),
     file = file.path(datadir, 'data.matrices.RData'))
```

### Specify the gene-gRNA group pairs to test for association

We include the genes used for GSFA in this analysis
```{r select-genes}
# Normalized and scaled data used for GSFA, the rownames of which are the 6k genes used for GSFA
scaled_gene_matrix <- TCells_data@assays$RNA@scale.data
dim(scaled_gene_matrix)
selected_gene_ids <- rownames(scaled_gene_matrix)

gRNA_groups <- colnames(metadata[, 4:24])
pairs <- expand.grid(selected_gene_ids, gRNA_groups)
gene_gRNA_group_pairs <- data.frame(gene_id = pairs$Var1, gRNA_group = pairs$Var2, pair_type = "candidate")
gene_gRNA_group_pairs[gene_gRNA_group_pairs$gRNA_group == "NonTarget", "pair_type"] <- "negative_control"
table(gene_gRNA_group_pairs$pair_type)
table(gene_gRNA_group_pairs$gRNA_group)
dim(gene_gRNA_group_pairs)
```

```{r, eval=FALSE}
saveRDS(gene_gRNA_group_pairs, file.path(datadir, "gene.gRNA.group.pairs.rds"))
```

### Run SCEPTRE

#### Stimulated T cells
```{r load-stimulated-input-data}
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data'
load(file.path(datadir, "data.matrices.RData"))
gene_gRNA_group_pairs <- readRDS(file.path(datadir, "gene.gRNA.group.pairs.rds"))

cat(sprintf('Dimenstion of gene expression matrix: %d rows %d columns.\n', nrow(gene_matrix), ncol(gene_matrix)))
cat(sprintf('Dimenstion of combined perturbation matrix: %d rows %d columns.\n', nrow(combined_perturbation_matrix), ncol(combined_perturbation_matrix)))
cat(sprintf('Dimenstion of covariate matrix: %d rows %d columns.\n', nrow(covariate_matrix), ncol(covariate_matrix)))
cat(sprintf('Dimenstion of gene gRNA-group pairs: %d rows %d columns.\n', nrow(gene_gRNA_group_pairs), ncol(gene_gRNA_group_pairs)))

table(gene_gRNA_group_pairs$pair_type)
```

```{r run-sceptre-stimulated-data, eval=FALSE}
result <- run_sceptre_high_moi(gene_matrix = gene_matrix,
                               combined_perturbation_matrix = combined_perturbation_matrix,
                               covariate_matrix = covariate_matrix,
                               gene_gRNA_group_pairs = gene_gRNA_group_pairs,
                               side = "both",
                               storage_dir = outdir, 
                               full_output = FALSE)

outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data/sceptre_output'
saveRDS(result, file.path(outdir, 'sceptre.result.rds'))
```

#### Unstimulated T cells
```{r load-unstimulated-input-data}
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/unsimulated_data'
load(file.path(datadir, "data.matrices.RData"))
gene_gRNA_group_pairs <- readRDS(file.path(datadir, "gene.gRNA.group.pairs.rds"))

cat(sprintf('Dimenstion of gene expression matrix: %d rows %d columns.\n', nrow(gene_matrix), ncol(gene_matrix)))
cat(sprintf('Dimenstion of combined perturbation matrix: %d rows %d columns.\n', nrow(combined_perturbation_matrix), ncol(combined_perturbation_matrix)))
cat(sprintf('Dimenstion of covariate matrix: %d rows %d columns.\n', nrow(covariate_matrix), ncol(covariate_matrix)))
cat(sprintf('Dimenstion of gene gRNA-group pairs: %d rows %d columns.\n', nrow(gene_gRNA_group_pairs), ncol(gene_gRNA_group_pairs)))

table(gene_gRNA_group_pairs$pair_type)
```

```{r run-sceptre-unstimulated-data, eval=FALSE}
result <- run_sceptre_high_moi(gene_matrix = gene_matrix,
                               combined_perturbation_matrix = combined_perturbation_matrix,
                               covariate_matrix = covariate_matrix,
                               gene_gRNA_group_pairs = gene_gRNA_group_pairs,
                               side = "both",
                               storage_dir = outdir, 
                               full_output = FALSE)

outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/unsimulated_data/sceptre_output'
saveRDS(result, file.path(outdir, 'sceptre.result.rds'))
```

### Analyze the results

#### Stimulated T cells

```{r load-stimulated-result}
outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data/sceptre_output'
simulated_result <- readRDS(file.path(outdir, 'sceptre.result.rds'))
head(simulated_result, 10)
```

Negative control pairs
```{r neg-control-stimulated}
neg_control_p_vals <- simulated_result %>% filter(pair_type == "negative_control") %>% pull(p_value)
qq_plot <- make_qq_plot(neg_control_p_vals)
plot(qq_plot)
```

Candidate pairs
We extract the p-values corresponding to the candidate pairs and apply a Benjamini-Hochberg (BH) correction to adjust for multiple testing.
```{r fdr-stimulated}
simulated_candidate_pair_results <- simulated_result %>% filter(pair_type == "candidate")
simulated_candidate_pair_results_p_adj <- simulated_candidate_pair_results %>%
  mutate(p_val_adj = p.adjust(p_value, method = "BH"))
head(simulated_candidate_pair_results_p_adj)
```

We call pairs with an adjusted p-value of less or equal than 0.1 significant; the discovery set (i.e., the set of significant pairs) has a false discovery rate (FDR) of 10%.
```{r select-sig-pairs-stimulated}
simulated_discovery_set <- simulated_candidate_pair_results_p_adj %>% filter(p_val_adj <= 0.1)
head(simulated_discovery_set)
```

```{r}
saveRDS(simulated_candidate_pair_results_p_adj, file.path(outdir, 'sceptre.candidate.pair.results.rds'))
saveRDS(simulated_discovery_set, file.path(outdir, 'sceptre.discovery.set.results.rds'))
```

#### Unstimulated T cells

```{r load-unstimulated-result}
outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/unsimulated_data/sceptre_output'
unsimulated_result <- readRDS(file.path(outdir, 'sceptre.result.rds'))
head(unsimulated_result, 10)
```

Negative control pairs
```{r neg-control-unstimulated}
neg_control_p_vals <- unsimulated_result %>% filter(pair_type == "negative_control") %>% pull(p_value)
qq_plot <- make_qq_plot(neg_control_p_vals)
plot(qq_plot)
```

Candidate pairs
We extract the p-values corresponding to the candidate pairs and apply a Benjamini-Hochberg (BH) correction to adjust for multiple testing.
```{r fdr-unstimulated}
unsimulated_candidate_pair_results <- unsimulated_result %>% filter(pair_type == "candidate")
unsimulated_candidate_pair_results_p_adj <- unsimulated_candidate_pair_results %>%
  mutate(p_val_adj = p.adjust(p_value, method = "BH"))
head(unsimulated_candidate_pair_results_p_adj)
```

We call pairs with an adjusted p-value of less or equal than 0.1 significant; the discovery set (i.e., the set of significant pairs) has a false discovery rate (FDR) of 10%.
```{r select-sig-pairs-unstimulated}
unsimulated_discovery_set <- unsimulated_candidate_pair_results_p_adj %>% filter(p_val_adj <= 0.1)
head(unsimulated_discovery_set)
```

```{r}
saveRDS(unsimulated_candidate_pair_results_p_adj, file.path(outdir, 'sceptre.candidate.pair.results.rds'))
saveRDS(unsimulated_discovery_set, file.path(outdir, 'sceptre.discovery.set.results.rds'))
```
